Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2022-12-19 paper-1/manuscript.tex
Spaces around =

800 excitatory and 200 inhibitory neurons were placed randomly in a two-dimensional circular space with a radius of 4~mm, and formed synapses of weight $w=1$ with 50 randomly selected postsynaptic partners within a range of 2~mm (excitatory) or 1~mm (inhibitory), reflecting the notion that excitatory neurons project over longer distances, while inhibitory neurons mainly connect to their local neighborhood. Connectivity and weights remained fixed throughout. Stimulation sites were evenly distributed 2.5~mm from the center of the dish, and stimulation delivered a one-time increase in $g_e$ to the 10 closest neurons, sufficient to trigger 2-3 spikes each. See Figure \ref{fig1}b and c for a representative example of the resulting network structure. Networks were pseudo-randomly generated according to the above scheme and screened for minimal stimulus response: Networks where fewer than 500 neurons responded to stimulation at any of the five sites after full recovery were discarded. In total, 30 networks were used for the experiments described below.

@@

Based on these measures, we then asked why there were more spikes in dev trials. We first focused on the contribution of $V_{TA}$ alone in an effort to confirm that thresholds are lower in dev, and that this reduction in thresholds causes increased firing. We correlated the level of $V_{TA}$ at the start of target trials (i.e., at the left edge of the histograms in Figure \ref{fig4}b) with the average target trial response of each excitatory neuron in terms of spikes fired. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in dev than msc (t = -17.3, p = 2.57e-57, median -0.166 mV; one-sided t-test, n=800), and this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 3.1e-36). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: Average $V_{TA}$ was lower in dev than msc (t = -7.32, p = 1.58e-11, median = -0.074 mV; one-sided t-test, n=120), and the neuron-level correlation between the contrasts in $V_{TA}$ and target response was almost universally negative (t = -23.1, p = 3.72e-46, median = -0.357). This indicates that networks were less adapted during oddball sequences, and that this lack of adaptation played an important role in the increased response to deviant inputs.

@@

Correlating the non-target response and thresholds excitatory neurons (Figure \ref{fig5}c), we found that the reduced non-target response magnitude in dev (t = -7.61, p = 3.5e-12, median = -0.044; one-sided t-test, n=800) almost perfectly predicted the resulting reduction in threshold adaptation levels (Pearson's $\rho$ = 0.99, p = 0).
This finding held across networks and stimuli, as shown in Figure \ref{fig5}d, with a median correlation coefficient of 0.994 and a consistent reduction in the average non-target response between msc and dev sequences (t = -11.6, p = 1.46e-21, median = -0.058; one-sided t-test, n=120), clearly validating our notion that the non-target trial response is the primary driver of threshold adaptation.

>>>>

800 excitatory and 200 inhibitory neurons were placed randomly in a two-dimensional circular space with a radius of 4~mm, and formed synapses of weight $w = 1$ with 50 randomly selected postsynaptic partners within a range of 2~mm (excitatory) or 1~mm (inhibitory), reflecting the notion that excitatory neurons project over longer distances, while inhibitory neurons mainly connect to their local neighborhood. Connectivity and weights remained fixed throughout. Stimulation sites were evenly distributed 2.5~mm from the center of the dish, and stimulation delivered a one-time increase in $g_e$ to the 10 closest neurons, sufficient to trigger 2-3 spikes each. See Figure \ref{fig1}b and c for a representative example of the resulting network structure. Networks were pseudo-randomly generated according to the above scheme and screened for minimal stimulus response: Networks where fewer than 500 neurons responded to stimulation at any of the five sites after full recovery were discarded. In total, 30 networks were used for the experiments described below.

@@

Based on these measures, we then asked why there were more spikes in dev trials. We first focused on the contribution of $V_{TA}$ alone in an effort to confirm that thresholds are lower in dev, and that this reduction in thresholds causes increased firing. We correlated the level of $V_{TA}$ at the start of target trials (i.e., at the left edge of the histograms in Figure \ref{fig4}b) with the average target trial response of each excitatory neuron in terms of spikes fired. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in dev than msc (t = -17.3, p = 2.57e-57, median -0.166 mV; one-sided t-test, n = 800), and this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 3.1e-36). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: Average $V_{TA}$ was lower in dev than msc (t = -7.32, p = 1.58e-11, median = -0.074 mV; one-sided t-test, n = 120), and the neuron-level correlation between the contrasts in $V_{TA}$ and target response was almost universally negative (t = -23.1, p = 3.72e-46, median = -0.357). This indicates that networks were less adapted during oddball sequences, and that this lack of adaptation played an important role in the increased response to deviant inputs.

@@

Correlating the non-target response and thresholds excitatory neurons (Figure \ref{fig5}c), we found that the reduced non-target response magnitude in dev (t = -7.61, p = 3.5e-12, median = -0.044; one-sided t-test, n = 800) almost perfectly predicted the resulting reduction in threshold adaptation levels (Pearson's $\rho$ = 0.99, p = 0).
This finding held across networks and stimuli, as shown in Figure \ref{fig5}d, with a median correlation coefficient of 0.994 and a consistent reduction in the average non-target response between msc and dev sequences (t = -11.6, p = 1.46e-21, median = -0.058; one-sided t-test, n = 120), clearly validating our notion that the non-target trial response is the primary driver of threshold adaptation.
